---
title: "Multivariate Analysis"
subtitle: |
  | Final Project 
  | Data Science 1 with R (STAT 301-1)
author: Jessie Chen
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: false
  warning: false

from: markdown+emoji 
---

::: {.callout-tip icon="false"}
## Github Repo Link

[My Repo Link](https://github.com/stat301-1-2023-fall/final-project-1-Jchench)
:::

```{r}
# loading dataset:
library(tidyverse)
spotify_data <- read_rds("data/spotify_data_cleaned")
```

As identified in progress memo 2, I'm going to look at a multivariate analysis between `final_streams`, `year`, `dance`, `loud`, `primary_genre`, `key`, and `song_seconds`. I actually don't think `time_signature` would be worthwhile to look at since most songs seem to be in 4/4 so the sample size for the other two time signatures would be way too small to draw any accurate conclusions.

Let's look at dancebility first:
```{r}
# final streams and dance:

spotify_data |> 
  filter(!is.na(primary_genre)) |> 
  ggplot(aes(x = dance, y = final_streams)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ primary_genre, scales = "free")
```






