---
title: "Progress Memo 1"
subtitle: |
  | Final Project 
  | Data Science 1 with R (STAT 301-1)
author: Jessie Chen
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: true
  warning: false

from: markdown+emoji 
---

::: {.callout-tip icon=false}

## Github Repo Link

[My Repo Link](https://github.com/stat301-1-2023-fall/final-project-1-Jchench)

```{r}
# Loading packages:
library(tidyverse)
library(naniar)
```

:::

## Data source

```{r}
# Loading data:

spotify_data_top <- 
  read_csv("data/ATop 1990-2022.csv")

spotify_streams <- 
  read_csv("data/spotify_streams.csv")

```

The `spotify_data_top` dataset I am using collected Spotify-made playlists that include between 50-130 tracks of the most popular songs each year from 1990-2022. There is information on each song's artist, album release date, genre, and other characteristics (i.e. BPM, key, and time signature). [This data was taken from Kaggle, courtesy of user KRISTOFFER.](https://www.kaggle.com/datasets/ztofff/top-1990-2022)

The `spotify_streams` dataset I am using collected data on 10000 Spotify songs and information on their number of streams. [This data was taken from Kaggle, courtesy of user RAKKESH ARAVIND G.](https://www.kaggle.com/datasets/rakkesharv/spotify-top-10000-streamed-songs)

I anticipate joining these data sets together and one way I could match them would be through the song name and artist name variables, which are listed differently in each dataset. Thus, I would have to rename these columns later so that they are consistent.

## Why this data

I chose to use these datasets because I think it would be interesting to look at the songs that were considered "popular" throughout the years form 1990-2022. The `spotify_data_top` has a lot of comprehensive information on the musical elements of each individual song (e.g. dancebility, genres, time signature, and key), and I think it could be useful to analyze if there are any correlations among these elements and "popularity". The `spotify_streams` dataset provides a more concrete, quantitative measure of "popularity" through collecting the number of streams (or how many times the song is played on the platform) for each song. Thus, I think it would be helpful to merge these two datasets together to look at the number of streams as an indicator of popularity.

## Data quality & complexity check

**Looking briefly at the data:**
```{r}
glimpse(spotify_data_top)

glimpse(spotify_streams)
```

The datasets seem to be sufficiently complex. `spotify_data_top` has 2610 observations and 26 variables. There are 12 categorical, 12 numeric, and 2 date and time-related variables. `spotify_streams` has 11084 rows and 9 variables. There are 3 categorical and 6 numeric variables. 

**Checking for missingness in `spotify_data_top`:**
```{r}
gg_miss_var(spotify_data_top)

miss_var_summary(spotify_data_top) |> 
  filter(n_miss != 0)
```
There don't seem to be too many variables with a large number of missing values. `song` and `artist` don't seem to have missing values, which is good because it will be easier to merge with the second dataset. However, many `genre` values seem to be missing though, which is something to keep in mind.

**Checking for missingness in `spotify_streams`:**
```{r}
gg_miss_var(spotify_streams)

miss_var_summary(spotify_streams) |> 
  filter(n_miss != 0)
```
This dataset only has one variable with missing values and it is `Song Name`. Unfortunately, this will make things a little bit more difficult going forwards because it won't be as easy to merge by song and artist name.

## Potential data issues

One potential issue I'm going to face in terms of merging is having to remove/ work around missing values. Since I'm thinking of merging by song and artist name, it will be a bit hard since a couple of song names in the `spotify_streams` dataset is missing. Another challenge I will face will be regarding the `date` and `time` variables in the `spotify_data_top` dataset. I might have to mutate the `date` column so that it shows the year, month, day separately. I may also need to find a better way to represent `time` numerically (perhaps convert to seconds). I also need to rename/ infer descriptions for a lot of the columns because the variable names are not quite clear in displaying what they're representing. For example, the `time` variable actually represents the length of the song in minutes and seconds. Another issue is that the `spotify_data_top` dataset does not have any information on songs from 2021, ^[the Kaggle user who provided this dataset could not find any information from 2021] which will impact my analysis going forward if I decide to look at songs across time. Overall, the datasets seem relatively clean.

## Misc

**Potential timeline:** I'm hoping to have my data fully cleaned and organized by the end of next week so I can start doing some initial exploration.
