---
title: "Exploring Data Songs in Spotify"
subtitle: |
  | Final Project Report
  | Data Science 1 with R (STAT 301-1)
author: Jessie Chen
date: today

format:
  html:
    toc: true
    embed-resources: true
    
execute:
  echo: false
  warning: false

from: markdown+emoji 
---

::: {.callout-tip icon=false}

## Github Repo Link

[My Repo Link](https://github.com/stat301-1-2023-fall/final-project-1-Jchench)

```{r}
# Loading packages and data:
library(tidyverse)
spotify_data <- read_rds("data/spotify_data_cleaned")
spotify_data_codebook <- read_csv("data/spotify_data_codebook.csv")
```

:::

## Introduction

I'm interested in looking at data on music to determine what musical elements make songs more popular and how these elements have changed over time. For this project, I gathered data from the music streaming-platform Spotify to look at some of the most popular songs among users and particular elements that might have made those songs popular. Some of the initial research questions I have are: What are the musical elements that make a song popular among listeners? Are there trends and patterns in musical elements and song popularity that have changed over time?

The `spotify_data_top` dataset I am using collected Spotify-made playlists that include between 50-130 tracks of the most popular songs each year from 1990-2022. There is information on each song's artist, album release date, genre, and other characteristics (i.e. BPM, key, and time signature). [This data was taken from Kaggle, courtesy of user KRISTOFFER.](https://www.kaggle.com/datasets/ztofff/top-1990-2022) The `spotify_streams` dataset I am using collected data on 10000 Spotify songs and information on their number of streams. [This data was taken from Kaggle, courtesy of user RAKKESH ARAVIND G.](https://www.kaggle.com/datasets/rakkesharv/spotify-top-10000-streamed-songs)


## Data overview & quality

`spotify_data_top` has 2610 observations and 26 variables. There are 12 categorical, 12 numeric, and 2 date and time-related variables. `spotify_streams` has 11084 rows and 9 variables. There are 3 categorical and 6 numeric variables.

In terms of potential issues, both the datasets doesn't seem to have too many variables with a large number of missing values. However, many `genre` values seem to be missing though, which is something to keep in mind. The biggest issue that I ran into when trying to join the two datasets was in regards to the keys that I decided to join by. I wanted to join the two data sets by artist and song names, however, this became hard to do because the datasets seemed to have multiple versions of the same songs by the same artists. I was able to overcome this issue by filtering for distinct values and summing the total number of streams. In my data cleaning, I also did a couple of variable conversions so that the dataset would be easier to work with.

## Explorations

**Addressing the first research question I proposed: What are the musical elements that make a song popular among listeners?**

I decided to define "popularity" of songs based on how many total streams a song garnered on Spotify. Here is the distribution for this variable:

```{r}
spotify_data |> 
  ggplot(aes(x = final_streams)) +
  geom_histogram() +
  labs(title = "Distribution of the total number of streams for songs",
       x = "total number of streams",
       y = NULL) +
  scale_x_continuous(labels = scales::label_number())
```
As we can see here, the distribution of the total number of streams is unimodal and significantly skewed to the right, meaning that most of the songs in this dataset have less than 100 million total streams. There seem to be a lot of outliers after 400 million total streams so we may want to just ignore these values for the next couple of explorations.

**Looking at the most important numeric variable:**

To begin with I looked at the correlation coefficient between all the numeric variables and the total number of streams:

```{r}
spotify_numeric <- 
  spotify_data |> 
  select(where(is.numeric))

spotify_numeric |> 
  cor(use = "pairwise.complete.obs") |> 
  as_tibble() |> 
  select(final_streams) |> 
  mutate(variables = colnames(spotify_numeric), .before = final_streams) |> 
  arrange(desc(final_streams)) |> 
  knitr::kable(digits = 2,
               col.names = c("Variables", "Correlation Coefficient"),
               caption = "Numeric Variables Correlated with Total Number of Streams")
```
Other than the obvious variable of `final_stream` (the total number of streams) and the non-musical element-related variable of `year`, this shows that the "danceability" of a song (represented in percentage) is the most correlated with total stream numbers (correlation coefficient of 0.15). 

Thus, I decide to looks more closely at this variable through a univariate analysis:

```{r}
spotify_data |> 
  ggplot(aes(x = dance)) +
  geom_histogram(bins = 40) +
  labs(title = "Distribution of danceability for songs",
       x = "danceability in %",
       y = NULL)
```
We can see that the distribution of danceability seems to have more or less a unimodal normal distribution. There might be a couple of song outliers below the 25% mark for danceability, so I may ignore them for explorations going forward.

Next, I decided to look at a bivariate analysis for a song's danceability and the total number of streams:
```{r}
spotify_data |> 
  ggplot(aes(x = dance, y = final_streams)) +
  geom_point(alpha = 0.6) +
  geom_jitter(width = 3, height = 2) +
  geom_smooth(method = "lm") +
  labs(title = "Total Number of Spotify Streams vs. Danceability",
       x = "dancebility (measured in %)",
       y = "total streams") +
  scale_y_continuous(labels = scales::label_number()) +
  coord_cartesian(y = c(0, 200000000),
                  x = c(25, 100))
```
We can see here that there is a relatively weak positive correlation between the total number of Spotify streams for a song and its danceability. Nonetheless, this means that songs that are more "danceable" tend to be played somewhat more often by Spotify users, and thus are more popular.

Lastly, I looked at a multivariate analysis of danceability, the total number of streams, and genre:

```{r}
spotify_data |> 
  filter(!is.na(primary_genre)) |> 
  ggplot(aes(x = dance, y = final_streams)) +
  geom_point(alpha = 0.2) +
  geom_jitter(width = 3, height = 2) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~ primary_genre, scales = "free") +
  scale_y_continuous(labels = scales::label_number()) +
  labs(title = "Total Number of Spotify Streams vs. Danceability For Each Genre",
       x = "dancebility (measured in %)",
       y = "total streams") +
  coord_cartesian(y = c(0, 200000000),
                  x = c(25, 95))
```
The relationship between danceability and total number of streams changes when we look at different genres. For Metal, country, folk, easy listening, and electronic, the relationship is negative, whereas for all the other genres, the relationship is positive. This means that listeners have different preferences for how "danceable" a song is depending on what genre they are listening to.

**Looking at the most important categorical variable:**

An important categorical variable I decided to explore when considering this research question was key signature. In particular, I decided to only focus on differentiating between major and minor keys to see if either feature would correlate with more streams. I did a bivariate analysis between a song's key signature and total number of streams to see if there are any noticeable patterns:

```{r}
spotify_data |> 
  ggplot(aes(x = major_minor, y = final_streams)) +
  geom_boxplot() +
  labs(title = "Distribution of the Total Number of Streams by Key",
       subtitle = "For Songs in a Major Key vs. Minor Key",
       x = "Major or Minor",
       y = "Total number of streams") +
  scale_y_continuous(labels = scales::label_number()) +
  coord_cartesian(ylim = c(0, 400000000))
```
It looks like the medians for the total number of streams for songs in either major or minor keys seem to be around the same. However, the range and interquartile range is greater for songs that are in the minor key. This means that songs in the minor key have a greater change of garnering more stream numbers.

Then, I did a multivariate analysis for the major/ minor key signature, total number of streams, and genres:

```{r}
spotify_data |> 
  filter(!is.na(primary_genre)) |> 
  ggplot(aes(x = primary_genre)) +
  geom_bar(aes(fill = major_minor)) +
  facet_wrap(~ stream_breaks, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Songs in Major or Minor Keys For Each Genre",
       subtitle = "Separated by number of streams",
       y = NULL,
       x = "Major or Minor Key")
```
As we can see here, the number of songs for each genre that are in either a major or minor key differs among different total stream numbers. Songs that have less than 250 million streams are typically in a major key. However, this kind of changes for songs that have less than 500 million streams as more hip hop songs in this category seem to be in the minor key. Lastly, when we look at songs that have less than 1 billion streams, we can see that for pop songs, more of them are in minor keys. However, for hip hop songs, more songs in this category seem to be in minor keys. Thus, there generally seems to be a pattern for songs in the hip hop and pop genres respectively. For the hip hop genre, songs that have more stream numbers seem to more likely be in major keys. For pop songs, on the other hand, songs that have more stream numbers seem to more likely be in minor keys.

**Addressing the second research question I proposed: Are there trends and patterns in musical elements and song popularity that have changed over time?**

I decided to first look at the distribution of years for when songs are released:

```{r}
spotify_data |> 
  ggplot(aes(x = year)) +
  geom_bar() +
  labs(title = "Distribution of Song Release Years",
       x = "Year",
       y = NULL)
```
We can see that the distirbution is left skewed and it seems like there are a lot of songs in this dataset from 2014, 2016, and 2017. There might be outliers at around 1980, so we might ignore this portion for the next couple of explorations.

**Looking at the most important numeric variable:**

To begin with I looked at the correlation coefficient between all the numeric variables and the song release year:

```{r}
spotify_numeric |> 
  cor(use = "pairwise.complete.obs") |> 
  as_tibble() |> 
  select(year) |> 
  mutate(variable = colnames(spotify_numeric), .before = year) |> 
  arrange((year)) |> 
  knitr::kable(digits = 2,
               col.names = c("Variables", "Correlation Coefficient"),
               caption = "Numeric Variables Correlated with Year")
```
We can see from the table that song duration (measure in seconds) is most closely correlated with the year in which a song is released (correlation coefficient of -0.43). 

Thus, I decided to conduct a bivariate analysis between song duration and year:

```{r}
spotify_data |> 
  ggplot(aes(x = year, y = song_seconds)) +
  geom_point(alpha = 0.5) +
  geom_jitter(width = 3, height = 1) +
  geom_smooth(method = "lm") +
  labs(title = "Song Duration vs. Year",
       x = "Year",
       y = "Song Duration (in seconds)")
```
The bivariate analysis supports the moderately strong, negative correlation between year and song duration. This shows that most songs, in general, have been getting shorter and shorter in length each year.

Next, let's look at a multivariate analysis between year, song duration, and total number of streams:

```{r}
spotify_data |> 
  ggplot(aes(x = year, y = song_seconds, color = stream_breaks)) +
  geom_point(alpha = 0.5) +
  geom_jitter(width = 3, height = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Song Duration vs. Year",
       subtitle = "Categorized by total stream numbers",
       x = "Year",
       y = "Song Duration (in seconds)",
       color = "Total Number of Streams")
```
As we can see here, songs duration seems to have decreased over the years for all total stream numbers. This means that songs, in general, regardless of popularity, have been getting shorter and shorter.

**Looking at the most important categorical variable:**

An important categorical variable I decided to explore when considering this research question is genre. 

```{r}
spotify_data |> 
  filter(!is.na(primary_genre)) |> 
  ggplot(aes(x = year)) +
  geom_bar(aes(fill = primary_genre)) +
  labs(title = "Distribution of Songs Released Each Year By Genre",
       x = "Year",
       y = NULL, 
       fill = "Genre")
```
As we can see from the bar graph, pop songs seem to constitute a significant portion of the songs released each year. Notably, rock songs used to be released more often in the 90s - late 2010s, but seem to have become less common in the 2020s. Lastly, more songs in the Latin genre have been released in recent years.

Next, let's look at a multivariate analysis between genre, year, and total number of streams:

```{r}
spotify_data |> 
  filter(!is.na(primary_genre)) |> 
  ggplot(aes(x = year)) +
  geom_density(aes(fill = primary_genre), alpha = 0.5) +
  facet_wrap(~ stream_breaks, scales = "free") + 
  labs(title = "Distribution of Songs Released Each Year By Genre",
       subtitle = "Categorized by total stream numbers",
       x = "Year",
       y = NULL, 
       fill = "Genre")
```
As we can see from the multivariate analysis, the distribution for song genres in each category of total stream numbers differs a lot. For the songs that have less than 250 million streams, songs in the Latin genre have mainly been released in recent years, and songs in the easy listening genre have been mainly released in the period around 2015. For songs that have less than 500 million streams, the distribution of songs in the pop and hip hop genres seem to peak around the 2017 period. Lastly, for songs that have less than 1 billion streams, the distribution of pop songs has stayed fairly constant throughout the late 90s - early 2020s period. However, the distribution for songs in the hip hop genre seems to peak around 2019. This multivariate analysis generally suggests that there has been a slight shift in the number of songs released from each genre throughout the years, changing from more Rock song releases to Latin and Hip Hop song releases. The most popular songs also seem to only come from a couple of genres such as hip hop and pop.

## Conclusion

Here are song of the main insights and conclusions that I have generated for my research questions based on my exploratory data analysis:

**What are the musical elements that make a song popular among listeners?**

I found that danceability and key signature (in major or minor) are important elements that might influence whether or not a song is popular among listeners. Although danceability and the total number of streams a song garners is mostly positive (the more "danceable" a song is, the more times it is played), the relationship between danceability and total number of streams changes when we look at different genres. For example, in the metal, country, folk, easy listening, and electronic genres, the relationship is negative. This means that listeners have different preferences for how "danceable" a song is depending on what genre they are listening to. In terms of key signature, I found that generally songs in the minor key have a greater possibility of having more total stream numbers. However this changes slightly when looking at songs from different genres. In particular, songs in the hip hop genre that have more stream numbers seem to more likely be in major keys and pop songs, on the other hand, that have more stream numbers seem to more likely be in minor keys.

**Are there trends and patterns in musical elements and song popularity that have changed over time?**

I found that song duration (in seconds) has seemed to decreased over the years, and this is mostly true for songs of any number of total streams. I kind of expected this finding because do feel like media content in general has gotten shorter and shorter recently due to development in social media and other technologies. I was a little bit more surprised when looking at the genres of song releases each year as I didn't really expect the number of songs in the Latin category to have increased in these past few years. I did expect hip hop and pop music genres to constitute most of the song releases, since I did notice that a large number of the songs in this dataset fall under these categories.

## References

**Data from:**

G, Rakkesh Aravind. “Spotify Top 10000 Streamed Songs.” Kaggle, 2 Jan. 2023, www.kaggle.com/datasets/rakkesharv/spotify-top-10000-streamed-songs. 

Kristoffer. “Spotify Top Tracks 1990-2022.” Kaggle, 27 Aug. 2023, www.kaggle.com/datasets/ztofff/top-1990-2022. 


